---
title: "Darwin Core mapping"
subtitle: "For: Checklist of alien species of the Scheldt estuary"
author:
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
#  pdf_document:
#    df_print: kable
#    number_sections: yes
#    toc: yes
#    toc_depth: 3
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/alien-scheldt-checklist/blob/master/src/dwc_mapping.Rmd).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

Load libraries:

```{r message = FALSE}
library(tidyverse)      # To do data science
library(googlesheets4)  # To import data from google sheets
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
```

# Read source data

Create a data frame `input_data` from the source data:

```{r}
input_data <- read_sheet("https://docs.google.com/spreadsheets/d/1V_rQhyz_AjszobCMxBe-xFWI3kUNLrYzMeqyUNt3Vc4/edit#gid=0", col_types = "c")
```

Preview data:

```{r}
input_data %>% head(n = 5)
```

Save as raw data file:

```{r}
input_data %>% write.csv(here::here("data", "raw", "input_data.csv"), row.names = FALSE)
```

# Preprocessing

## Tidy data

Clean dataset names:

```{r}
input_data %<>% clean_names()
```

## Generate Taxon ID

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom (that will remain the same as long as scientific name and kingdom remain the same):

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "alien-scheldt-checklist",
  "taxon",
  vdigest(paste(scientific_name, kingdom), algo = "md5"),
  sep = ":"
))
```

Preview data:

```{r}
input_data %>% head()
```

# Darwin Core mapping

## Create taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
taxon <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/") 
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "INBO")
```

### datasetID

```{r}
taxon %<>% mutate(dwc_datasetID = "") 
```

### institutionCode

```{r}
taxon %<>% mutate(dwc_institutionCode = "INBO")
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "Checklist of alien species of the Scheldt estuary, Belgium")
```

### taxonID

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = scientific_name)
```

### kingdom

```{r}
taxon %<>% mutate(dwc_kingdom = kingdom)
```

### phylum

```{r}
taxon %<>% mutate(dwc_phylum = phylum)
```

### class

```{r}
taxon %<>% mutate(dwc_class = class)
```

### order

```{r}
taxon %<>% mutate(dwc_order = order)
```

### family

```{r}
taxon %<>% mutate(dwc_family = family)
```

### genus

```{r}
taxon %<>% mutate(dwc_genus = genus)
```

### taxonRank

```{r}
taxon %<>% mutate(dwc_taxonRank = taxon_rank)
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(dwc_nomenclaturalCode = "ICZN")
```

## Create distribution extension

```{r}
distribution <- input_data
```

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(dwc_taxonID = taxon_id)
```

### locationID

```{r}
distribution %>% group_by(country_code) %>%
  count()
```

Map values to [ISO 3166](https://www.iso.org/iso-3166-country-codes.html) country codes

```{r}
distribution %<>% mutate(dwc_locationID = "ISO_3166-2:BE")
```

### locality

Use the country name to map `locality`:

```{r}
distribution %<>% mutate(dwc_locality = "Belgium")
```

### countryCode

```{r}
distribution %<>% mutate(dwc_countryCode = country_code) 
```

### occurrenceStatus

Inspect values:

```{r}
distribution %>%
  group_by(occurrence_status) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_occurrenceStatus = occurrence_status) 
```

### establishmentMeans

```{r}
distribution %<>% mutate(dwc_establishmentMeans = "introduced") 
```

### eventDate

Inspect `date_first_observation`:

```{r}
distribution %>%
  group_by(date_first_observation) %>%
  count() 
```

Remove `?` or `<`

```{r}
distribution %<>% mutate(date_first_observation = str_remove_all(date_first_observation,"[?|<|>]"))
```

Inspect `date_last_observation`:

```{r}
distribution %>%
  group_by(date_last_observation) %>%
  count() 
```

Combine `start_year` and `end_year` in an ranged `Date` (ISO 8601 format). If any those two dates is empty or the same, we use a single year, as a statement when it was seen once (either as a first record or a most recent record):

```{r}
distribution %<>% mutate(dwc_eventDate = case_when(
  is.na(date_first_observation) & is.na(date_last_observation) ~ "",
  is.na(date_first_observation) & !is.na(date_last_observation) ~ date_last_observation,
  !is.na(date_first_observation) & is.na(date_last_observation) ~ date_first_observation,
  date_first_observation == date_last_observation ~ date_last_observation,
  TRUE ~ paste(date_first_observation, date_last_observation, sep = "/")
))
```

### occurrenceRemarks

Inspect values:

```{r}
distribution %>%
  group_by(remarks) %>%
  count()
```

Map values:

```{r}
distribution %<>% mutate(dwc_occurrenceRemarks = remarks) 
```

## Create species profile extension

In this extension will express broad habitat characteristics (e.g. `isTerrestrial`) of the species.

```{r start_species_profile}
species_profile <- input_data
```

Habitat information can be found in `realm`, which describes whether a species is found in freshwater, terrestrial or both (terrestrial/freshwater) habitats.

Show unique values:

```{r}
species_profile %>%
  group_by(realm) %>%
  count()
```

We map this information to `isFreshwater` and `isMarine` in the [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml):

### taxonID

```{r}
species_profile %<>% mutate(dwc_taxonID = taxon_id)
```

### isMarine

```{r}
species_profile %<>% mutate(dwc_isMarine = "TRUE")
```

### isFreshwater

```{r}
species_profile %<>% mutate(dwc_isFreshwater = "TRUE")
```

### isTerrestrial

```{r}
species_profile %<>% mutate(dwc_isTerrestrial = "FALSE")
```

Show mapped values:

```{r}
species_profile %>%
  select(realm, dwc_isMarine, dwc_isFreshwater, dwc_isTerrestrial) %>%
  group_by_all() %>%
  summarize(records = n())
```

## Create description extension

In the description extension we want to include several important characteristics (hereafter refered to as descriptors) about the species:

* Native range
* Degree of establishment

The structure of the description extension is slightly different from the other core/extension files: information for a specific taxon (linked to taxonID) is provided in **multipele** lines within the csv file: one line per taxon per descriptor. In this way, we are able to include multipele descriptors for each species. 

For each descriptor, we create a separate dataframe to process the specific information. We always specify *which descriptor* we map (`type` column) and its *specific content* (`description` column). After the mapping of these Darwin Core terms `type` and `value`, we merge the dataframes to generate one single description extension. We then continue the mapping process by adding the other Darwin Core terms (which content is independent of the type of descriptor, such as `language`).

### Native range

Create separate dataframe:

```{r start_native_range}
native_range <- input_data
```

Show unique values:

```{r}
native_range %>%
  distinct(native_range) %>%
  arrange(native_range)
```

`native_range` contains multiple values (currently not more than 3), so we separate it in 3 columns:

```{r}
native_range %<>% separate(native_range, 
  into = c("native_range_1", "native_range_2", "native_range_3"),
  sep = "\\s\\|\\s")
```

Gather in a `key` and `value` column:

```{r}
native_range %<>% gather(
  key, value,
  native_range_1, native_range_2, native_range_3,
  na.rm = TRUE, # Also removes records for which there is no native_range_1
  convert = FALSE
)
```

Sumarize `value`

```{r}
native_range %>%
  group_by(value) %>%
  count()
```

Recode `probably the Americas` to `America`:

```{r}
native_range %<>% mutate(description = recode(value,
  "probably the Americas" = "America"))
```

Drop the `key` and `value` columns:

```{r}
native_range %<>% select(-key, -value)
```

Create a `type` field to indicate the type of description:

```{r}
native_range %<>% mutate(type = "native range")
```

### Degree of establishment

Here, we use the [invasion stage vocabulary from Blackburn et al. (2011)](http://doc.rero.ch/record/24725/files/bach_puf.pdf) to standardize this information.

```{r}
degree_of_establishment <- input_data
```

Summarize content

```{r}
degree_of_establishment %>%
  group_by(degree_of_establishment) %>%
  count()
```

Create `description`:

```{r}
degree_of_establishment %<>% mutate(description = degree_of_establishment)
```

Create `type`:

```{r}
degree_of_establishment %<>% mutate(type = "degree_of_establishment")
```

### Union description

Union native range and degree of establishment:

```{r start_description_ext}
description_ext <- bind_rows(native_range, degree_of_establishment)
```

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml):

#### taxonID

```{r}
description_ext %<>% mutate(dwc_taxonID = taxon_id)
```

#### description

```{r}
description_ext %<>% mutate(dwc_description = description)
```

#### type

```{r}
description_ext %<>% mutate(dwc_type = type)
```

#### language

```{r}
description_ext %<>% mutate(dwc_language = "en")
```

# Post-processing

Remove the original columns for each of the generated files:

```{r}
taxon %<>% select(starts_with("dwc_")) 
distribution %<>% select(starts_with("dwc_")) 
species_profile %<>% select(starts_with("dwc_")) 
description_ext %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
colnames(distribution) <- str_remove(colnames(distribution), "dwc_")
colnames(species_profile) <- str_remove(colnames(species_profile), "dwc_")
colnames(description_ext) <- str_remove(colnames(description_ext), "dwc_")
```

Sort on `taxonID` to group description information per taxon:

```{r}
taxon %<>% arrange(taxonID)
distribution %<>% arrange(taxonID)
species_profile %<>% arrange(taxonID)
description_ext %<>% arrange(taxonID)
```

Preview taxon core:

```{r}
taxon %>% head(10)
```

Preview distribution extension:

```{r}
distribution %>% head(10)
```

Preview species profile extension:

```{r}
species_profile %>% head(10)
```

Preview description extension:

```{r}
description_ext %>% head(10)
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description_ext, here("data", "processed", "description.csv"), na = "")
```


